\LARGE\textbf{Designing Studies and Evaluating Research Results:\\ Type M and Type S Errors for\\ Pearson Correlation Coefficient }

### 2020-06-10 Documents/Paper_main/Paper_main.Rnw
Corrige Giulia

The central and possibly most difficult point in this process is reasoning on what could be the plausible magnitude and direction of the effect of interest. This critical process, which is central also in traditional power analysis, is an opportunity for researchers to aggregate, formalize and incorporate prior information on the phenomenon under investigation \parencite{gelmanPowerCalculationsAssessing2014}. What is plausible can be determined on theoretical grounds, using expert knowledge elicitations techniques and consulting literature reviews and meta-analysis, always taking into account the presence of effect sizes inflation in the published literature \todo{(Anderson, 2019)}. The plausible effect size approximates the true effect, which is never known but can be thought as “that which would be observed in a hypothetical infinitely large sample” \parencite[p. 642]{gelmanPowerCalculationsAssessing2014}. For a more exhaustive description on plausible effect size, we refer the interested reader to \textcite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

@@

In the following sections we illustrate how to perform prospective and retrospective design analysis using some examples. We developed two R functions to perform design analysis for Pearson correlation, which are available at the page \url{https://osf.io/9q5fr/}. The function to perform a prospective design analysis is \texttt{pro\_r()}. It requires as input the plausible effect size (\texttt{rho}), the statistical power (\texttt{power}) the directionality of the test (\texttt{alternative}) which can be set as: “\texttt{two.sided}”, “\texttt{less}” or “\texttt{greate}r” . Type I error rate (\texttt{sig\_level}) is set as default at 5\% and can be changed by the user. The \texttt{pro\_r()} function returns the necessary sample size to achieve the desired statistical power, Type M error rate, the Type S error probability, and the critical value(s) above which a statistically significant result can be found. The function to perform retrospective design analysis is \texttt{retro\_r()}. It requires as input the plausible effect size, the sample size used in the study, and the directionality of the test that was performed. Also in this case, Type I error rate is set as default at 5\% and can be changed by the user.  The function \texttt{retro\_r()} returns the Type M error rate, the Type S error probability, and the critical value(s). For further details regarding the R functions refer to Appendix~\ref{sec:appendix_b}.

@@

<<Plot_winner, fig.pos="!h", fig.align="center", fig.height=3, fig.width=4.5, fig.cap="Winner's course. H$_0$ = Null Hypothesis, H$_1$ = Alternative Hypothesis. When sample size, directionality of the test and Type error probability are set, also the smallest effect size above which is possible to find a statistically significant result is set. In this case, the plausible effect size, $\\rho=.25$, lies in the region where it is not possible to reject H$_0$ (the region delimited by the two vertical lines). Thus, it is impossible to simultaneously find a statistically significant result and an effect close to the plausible one. In other words, a statistically significant effect must exaggerate the plausible effect size.", fig.scap="Winner's course">>=

@@

We can implement the prospective design analysis using the function \texttt{pro\_r()} which inputs and outputs are displayed in Figure~\ref{fig:pro_r}. About 125 participants are necessary to have 80 \% probability to detect an effect of at least $\rho=\pm .25$, if it actually exists. With this sample size, the Type S error is minimized and approximates zero. In this study design, the Type M error is 1.12 indicating that statistically significant results are on average exaggerated by 12\%. It is possible to notice that the critical values are $r=\pm .18$, further highlighting that our plausible effect size is actually included among those values that lead to the acceptance of the alternative hypothesis.

>>>>

The central and possibly most difficult point in this process is reasoning on what could be the plausible magnitude and direction of the effect of interest. This critical process, which is central also in traditional power analysis, is an opportunity for researchers to aggregate, formalize and incorporate prior information on the phenomenon under investigation \parencite{gelmanPowerCalculationsAssessing2014}. What is plausible can be determined on theoretical grounds, using expert knowledge elicitations techniques and consulting literature reviews and meta-analysis, always taking into account the presence of effect sizes inflation in the published literature \parencite{andersonBestOftForgotten2019}. The plausible effect size approximates the true effect, which is never known but can be thought as “that which would be observed in a hypothetical infinitely large sample” \parencite[p. 642]{gelmanPowerCalculationsAssessing2014}. For a more exhaustive description on plausible effect size, we refer the interested reader to \textcite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

@@

In the following sections we illustrate how to perform prospective and retrospective design analysis using some examples. We developed two R functions to perform design analysis for Pearson correlation, which are available at the page \url{https://osf.io/9q5fr/}. The function to perform a prospective design analysis is \texttt{pro\_r()}. It requires as input the plausible effect size (\texttt{rho}), the statistical power (\texttt{power}), the directionality of the test (\texttt{alternative}) which can be set as: “\texttt{two.sided}”, “\texttt{less}” or “\texttt{greate}r” . Type I error rate (\texttt{sig\_level}) is set as default at 5\% and can be changed by the user. The \texttt{pro\_r()} function returns the necessary sample size to achieve the desired statistical power, Type M error rate, the Type S error probability, and the critical value(s) above which a statistically significant result can be found. The function to perform retrospective design analysis is \texttt{retro\_r()}. It requires as input the plausible effect size, the sample size used in the study, and the directionality of the test that was performed. Also in this case, Type I error rate is set as default at 5\% and can be changed by the user.  The function \texttt{retro\_r()} returns the Type M error rate, the Type S error probability, and the critical value(s). For further details regarding the R functions refer to Appendix~\ref{sec:appendix_b}.

@@

<<Plot_winner, fig.pos="!h", fig.align="center", fig.height=3, fig.width=4.5, fig.cap="Winner's course. H$_0$ = Null Hypothesis, H$_1$ = Alternative Hypothesis. When sample size, directionality of the test and Type I error probability are set, also the smallest effect size above which is possible to find a statistically significant result is set. In this case, the plausible effect size, $\\rho=.25$, lies in the region where it is not possible to reject H$_0$ (the region delimited by the two vertical lines). Thus, it is impossible to simultaneously find a statistically significant result and an effect close to the plausible one. In other words, a statistically significant effect must exaggerate the plausible effect size.", fig.scap="Winner's course">>=

@@

We can implement the prospective design analysis using the function \texttt{pro\_r()} which inputs and outputs are displayed in Figure~\ref{fig:pro_r}. About 125 participants are necessary to have 80\% probability to detect an effect of at least $\rho=\pm .25$, if it actually exists. With this sample size, the Type S error is minimized and approximates zero. In this study design, the Type M error is 1.12 indicating that statistically significant results are on average exaggerated by 12\%. It is possible to notice that the critical values are $r=\pm .18$, further highlighting that our plausible effect size is actually included among those values that lead to the acceptance of the alternative hypothesis.
